\myTitle

### 2015-03-12 tex/response_to_refs_letter.tex

\end{itemize}

{\bf Referee Number 1 comments}

I have several important concerns over this paper:

\reply{I'll give brief replies to each of the following here in outline,
and will give more complete details in the following sections.}
\begin{enumerate}
\item The conceptual framework is inappropriate since FLOCK is not an MCMC algorithm
\reply{{\sc flock} is not an MCMC algorithm but it {\em is} an algorithm and an inference procedure and 
recognizing it as a special case of well known statistical methods is appropriate.}
\item There is only one dataset which is way too little to come to conclusions
\reply{We agree with the referee. We have undertaken a simulation study across a range of different $F_\mathrm{ST}$ values.}
\item There are basic flaws in the interpretation of the outputs of FLOCK, leading to invalid
comparisons relative to K values other than the estimate
\reply{We have eliminated issues of what can and can't be said about allocation of individuals  at a
certain value of $K$ by doing all the simulations and the analyses at a known value of $K$.}
\item Comparisons of estimates of K cannot be performed with sufficient rigor
\reply{We agree with the referee here.  In fact, we weren't entirely comfortable addressing the issue of estimating
$K$ in the previous version of the manuscript, lest it appear that we were implicitly encouraging it. 
Therefore, we now focus entirely on allocation of individuals.}
\item The speed comparisons are based on an ill-chosen processing unit
\reply{We implemented the {\sc flock} algorithm by modifying the C source code for {\sc structure} and
adding an R wrapper.  This allows a more direct comparison of the amount of computation involved.  If 
{\sc flock} were implemented as efficiently as {\sc structure} it would be much faster.  We report that now in the paper in a way that should be easily understood. }
\end{enumerate}

@@

{\bf 1) Conceptual framework}

It must be emphasized that FLOCK is not an MCMC based algorithm. It has no target function, no transition probabilities and does not move in small measured probabilistic steps. No amount of mathematical analysis will ever change this.

\reply{We didn't say {\sc flock} was an MCMC algorithm.  We noted that it was a limiting
 case of a simulated annealing algorithm.}

In fact, similar iterative methods have been around for a very long time (e.g. Newton's method for finding roots), much before the first MCMC algorithms were designed (1950?s). This family of methods should be analysed within a truly pertinent conceptual framework (attractors, basins of attraction, fixed points, orbits, etc). The book A First Course in Discrete Dynamical Systems (RA Holmgren, 1994) provides an excellent introduction to this topic. 
\reply{The referee is correct that iterative methods have been around for a long time.
In fact, rather than viewing it as special case of simulated annealing, one could just as well
note that {\sc flock} is a type of gradient descent-algorithm attempting to maximize the
likelihood of the data given a configuration of clusters.
Regardless of what name is given to it, since it is an algorithm
that (intentionally or not) yields solutions with 
high (and possibly maximal) likelihood, we find it quite clear that  
the conceptual framework of statistics is at least as ``truly pertinent,''
for purposes of analysis of the algorithm, as is
the conceptual framework of dynamical systems.
}

@@

{\bf 2) Data}

Performance comparisons between programs based on a single dataset are of little value. The fact that it is large and real does not compensate. For this purpose, one would expect that a large set of simulations spanning several parameter values with or without several real datasets would be used instead of a single empirical dataset. The advantage of simulations is, of course, that the true genetic structure of the data (essentially K) is known and so the relative performances may be compared on a solid basis. Here, unless I missed something, there is no independent (of the two programs) source of knowing with a reasonable degree of certainty the actual number of clusters represented within this collection of genotypes. In fact, there appears to be several candidate values for K with this particular dataset (see below).

\reply{We agree whole-heartedly and undertake numerous simulations of data like the real
steelhead data in the previous submission.  Having done this, and analyzed the results,
we can now address the comments below.}
Several (over)generalizations are based on this unique dataset:
\begin{itemize}
\item (p.9 ) FLOCK seems to find the same solution less consistently than does STRUCTURE
\reply{From our simulations using {\sc flockture} this is clearly true.  One need only look
at the variability present in the purple boxplots in Figures~2 and~3 to see that there is more
variability between {\sc flockture} analyses (50 runs of 20 iterations each) on the same data set 
than between {\sc structure} analyses. }
\item (p.14) The problem of getting caught in local modes is not unique to FLOCK...; however our results suggest it is a bigger problem for FLOCK
\reply{The simulation results bear this out, but we have reworded it to, ``This likely 
increases the number of
local peaks, leading {\sc flockture} to settle on different solutions in 
different runs. The problem presented by having multiple modes in the likelihood surface 
is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our simulations suggest this is a more  
prevalent problem for {\sc flockture}.''
}
\item (p.16) Given FLOCK'S tendency to converge to different solutions with large K and n...
\reply{See above.  Note that we don't vary $K$ now so we don't call it a problem ``with large $K$'' any
longer.}
\item (p.16) It is probably better to use STRUCTURE in such cases...
\reply{We no longer use this phrase specifically, but we do point out that {\sc flock} is not best-suited to high-gene-flow
species with very low $F_\mathrm{ST}$.}
\end{itemize}

@@

{\bf 3) Interpretation of FLOCK outputs (purpose of FLOCK, best run, stochasticity of plateau lengths)
Purpose of FLOCK}

I agree with the following statement (p. 11):
``While we take the stance that estimates of K made with any unsupervised clustering algorithm from data on real (non-idealized) populations should always be interpreted and used cautiously...''
I am fully aware that we are dealing with probabilistic statements here, not certainties. However, I am also convinced that the search for the number of genetic units (K) is a central issue. This is precisely what Flock intends to do, and it is also the main task that STRUCTURE was built for.

\reply{We disagree with the referee that estimating $K$ was the ``main task that {\sc structure} was
built for.''  I had the pleasure of visiting Jonathan Pritchard in Oxford shortly after the
{\sc structure} paper was published. At that time, the members of the Donnelly lab were not eagerly using
{\sc structure} to estimate $K$ from different samples; they were using {\sc structure} to correct
for population admixture when doing association tests in human genetics---our understanding
is {\em that} was the task that structure was originally built for.  Furthermore, when I contacted
Jonathan a few years later to ask him some details about how $\ln P(D)$ was calculated by {\sc structure},
he pointed out to me (while getting his notes out of his filing cabinet) that the method for estimating
$K$ was added into the paper to satisfy a referee who had requested it, and that Matthew Stephens 
(one of the co-authors) was not ever extremely pleased to have it in there.  Finally, I teach an annual
course with Matthew Stephens, and each year he explains to the students why estimating $K$ is so
difficult and why estimates of $K$ from {\sc structure} or any other procedure should be 
regarded with circumspection.  }

@@

{\em Misinterpretation of `best run'}

However, the authors are in fact proposing to analyse the output of FLOCK for each K as if it were potentially informative. This, in turn, lead to a misunderstanding of what `best run' means. For instance on p.12, they state:
``For all groups of runs for $K> 2$ no plateaus were observed...it is unclear in these cases what constitutes the `best run' as all plateau lengths are equal to one.''

The answer is simple: there is no best run. The term `best run' applies exclusively to the run with the longest plateau corresponding to the estimate of K. First the stopping rules are applied then the estimation rules. If the estimate is not undecided, the only best run to be considered is the one associated with the estimated K. If the estimate is undecided there is no best run. The documentation and also the flow chart in the latest article (Duchesne and Turgeon 2012) make that clear. In fact, outside the estimated K, no partitions from FLOCK should be considered. Therefore comparisons between clusterings from FLOCK and STRUCTURE for all K except the estimated one are just irrelevant and not valid.

\reply{This seems strange to us.  What happens in the situation where the
``true $K$'' is known and you just want to cluster individuals? If there are no plateaus
amongst the runs at that value of $K$ should one discard {\sc flock} and use
a different program altogether? In most of the simulated data sets we investigated, {\sc flock},
like {\sc flockture},
would not have seen any plateaus at all!---and yet it is still delivering useful clustering
inference that is competitive with, and at times better than, {\sc structure}.  We discuss
this rather restrictive guidance on using {\sc flock} in the Discussion.  }

### 2015-03-12 tex/main-body-text.tex
a few more late night changes.

  \begin{figure*}
\centering
\includegraphics[width=.9\linewidth]{images/Figures-Pat/uSatLoss.pdf}%uSatloss_BW.pdf is a greyscale version
  \caption{Zero-one loss for nine runs of {\sc flockture} and the admixture and correlated allele 
  frequency (A.C) model, the no-admixture and correlated allele frequency (NA.C) model, 
and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure}. Forty-eight 
datasets, composed of 2,000 individuals with 15 microsatellite genotypes comprising
5 populations of sizes 250, 250, 400, 550, and 550, were simulated to evaluate the performance
of both programs. }
  \label{fig:uSatLoss}
\end{figure*}

@@

Though we observed similar results
between the two programs at large values of $F_{ST}$, 
{\sc flockture}'s solutions consistently assigned fewer individuals correctly---and the
replicate solutions for individual data sets became more variable---at lower values of $F_{ST}$.
This is likely a consequence of the low population divergence which
makes it possible for many different allocations of individuals to
have roughly the same likelihood.
The likelihood surface defined on partitions of this space is then often
be multimodal---possessing numerous local peaks and troughs.
The peak that the {\sc flock} algorithm finds is entirely determined by the initial, random, allocation of 
individuals, as the algorithm proceeds deterministically after the initial random allocation.
Decreasing $F_{ST}$ leads to more homogenous 
populations whose multilocus genotypes have a smaller difference in 
likelihood of assigning to any given cluster. This likely increases the number of
local peaks, leading {\sc flockture} and {\sc flock} to settle on many different solutions in each run. The problem presented by having multiple modes in the likelihood surface 
is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our simulations suggest this is a more  
prevalent problem for {\sc flockture}.
{\sc flock} is not best-suited to species characterized by large
subpopulations and low divergence, as is the case with many 
marine fish.

>>>>

  \begin{figure*}
\centering
\includegraphics[width=.9\linewidth]{images/Figures-Pat/uSatLoss.pdf}%uSatloss_BW.pdf is a greyscale version
  \caption{Zero-one loss (lower values = better performance) for nine runs of {\sc flockture} and the admixture and correlated allele 
  frequency (A.C) model, the no-admixture and correlated allele frequency (NA.C) model, 
and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure}. Forty-eight 
datasets, composed of 2,000 individuals with 15 microsatellite genotypes comprising
5 populations of sizes 250, 250, 400, 550, and 550, were simulated to evaluate the performance
of both programs. }
  \label{fig:uSatLoss}
\end{figure*}

@@

Though we observed similar results
between the two programs at large values of $F_{ST}$, 
{\sc flockture}'s solutions consistently assigned fewer individuals correctly---and the
replicate solutions for individual data sets became more variable---at lower values of $F_{ST}$.
This is likely a consequence of the low population divergence which
makes it possible for many different allocations of individuals to
have roughly the same likelihood.
The likelihood surface defined on partitions of this space is then often
be multimodal---possessing numerous local peaks and troughs.
The peak that the {\sc flock} algorithm finds is entirely determined by the initial, random, allocation of 
individuals, as the algorithm proceeds deterministically after the initial random allocation.
Decreasing $F_{ST}$ leads to more homogenous 
populations whose multilocus genotypes have a smaller difference in 
likelihood of assigning to any given cluster. This likely increases the number of
local peaks, leading {\sc flockture} and {\sc flock} to settle on settle on different solutions in 
different runs. The problem presented by having multiple modes in the likelihood surface 
is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our simulations suggest this is a more  
prevalent problem for the {\sc flock} algorithm.
{\sc flock} is not best-suited to species characterized by large
subpopulations and low divergence, as is the case with many 
marine fish.
